---
title: "Preparing MERRA-2 Data to be Used for Examples in `listenr`"
date: "<br> Updated on `r format(Sys.time(), '%B %d, %Y')`"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

This notebook contains information and code used to generate the data `merra2` contained in `listenr`.

Load R packages:

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
```

Load data:

- These datasets were manually downloaded from the Goddard Earth Sciences Data and Information Services Center: [https://disc.gsfc.nasa.gov/](https://disc.gsfc.nasa.gov/). The files contain data from Modern-Era Retrospective Analysis for Research and Applications, Version 2 (MERRA-2). Specifically: 

  - `aod_raw`: contains vertically integrated monthly AOD values (obtained from [https://disc.gsfc.nasa.gov/datasets/M2TMNXAER_5.12.4/summary](https://disc.gsfc.nasa.gov/datasets/M2TMNXAER_5.12.4/summary)) 
  - `temp_strat_raw`: contains monthly stratospheric temperatures at 50 mb (obtained from [https://disc.gsfc.nasa.gov/datasets/M2IMNPASM_5.12.4/summary](https://disc.gsfc.nasa.gov/datasets/M2IMNPASM_5.12.4/summary)) 

- For more information on MERRA-2 see the following references: 
  - Gelaro, R., W. McCarty, M. J. Suarez, R. Todling,A. Molod, L. Takacs, C. A. Randles, A. Darmenov,M. Bosilovich, R. Reichle, K. Wargan, L. Coy, R. Cul-lather, C. Draper, S. Akella, V. Buchard, A. Conaty, A. M.da Silva, W. Gu, G.-K. Kim, R. Koster, R. Lucchesi,D. Merkova, J. E. Nielsen, G. Partyka, S. Pawson, W. Put-man, M. Rienecker, S. D. Schubert, M. Sienkiewicz,and B. Zhao, 2017: The modern-era retrospective anal-ysis for research and applications, version 2 (merra-2).Journal of Climate,30, 5419–5454, doi:10.1175/JCLI-D-16-0758.1.
  - Randles, C. A., A. M. da Silva, P. Colarco, A. Darmenov,R. Govindaraju, A. Smirnov, B. Holben, R. Ferrare,J. Hair, Y. Shinozuka, and C. Flynn, 2017: The merra-2aerosol reanalysis, 1980 onward. part i: System descrip-tion and data assimilation evaluation.Journal of Climate,30, 6823–6850, doi:10.1175/JCLI-D-16-0609.1.
  
```{r}
aod_raw = read.csv("merra2/tavgM_2d_aer_Nx/TOTEXTTAUall_48x24.csv")
temp_strat_raw = read.csv("merra2/instM_3d_asm_Np/MERRA2_3dasm_temperature_50mb_48x24.csv")
```

Clean up datasets in order to join:

- Remove the `lev` variable from the stratospheric data
- Rename variables with better labels
- Convert date to `lubridate` variable -- importance since character strings are different between datasets (e.g., `r aod_raw$date[1]` vs `r temp_strat_raw$date[1]`)
- Correct stratosphere day (I noticed that stratosphere day was 0 and should be 1)

```{r}
aod <- 
  aod_raw %>% 
  rename("aod" = "TOTEXTTAU") %>%
  mutate(date = as_date(date))

temp_strat <-
  temp_strat_raw %>% 
  select(-lev) %>% 
  rename("temp_strat" = `T`) %>%
  mutate(date = as_date(date)) %>%
  mutate(day = day(date))
```

Join the cleaned data and reorder columns:

```{r}
merra2_full <- 
  full_join(aod, temp_strat, by = c("lon", "lat", "date", "year", "month", "day")) %>%
  select(lon, lat, date, year, month, day, everything()) 
```

Preview part of the cleaned data

```{r}
head(merra2_full)
```

Plot of variables:

```{r fig.width = 14, fig.height = 6, echo = FALSE}
merra2_full %>%
  pivot_longer(cols = aod:temp_strat, names_to = "var") %>%
  group_by(date, var) %>%
  summarise(global_mean = mean(value, na.rm = TRUE),
            .groups = "drop") %>%
  ggplot(aes(x = date, y = global_mean)) +
  geom_line() +
  geom_point() +
  facet_grid(var ~ ., scales = "free_y", switch = "y") +
  theme_bw() +
  theme(strip.background = element_blank(),
        strip.placement = "outside") +
  labs(
    x = "Date",
    y = "",
    title = "Global Means of MERRA2 Variables"
  )
```

Assign training/testing data labels, extract necessary variables for ESN model, filter years, and make sure data are arranged appropriately: 

```{r}
merra2_train_test <-
  merra2_full %>%
  filter(year >= 1986 & year <= 1995) %>%
  mutate(data = ifelse(year > 1993, "test", "train")) %>%
  select(data, date, lon, lat, aod, temp_strat) %>%
  arrange(lon, lat, date)
```

Compute spatial means on training data:

```{r}
merra2_spatial_means_and_sds <- 
  merra2_train_test %>%
  filter(data == "train") %>%
  group_by(lon, lat) %>%
  summarise(
    aod_mean = mean(aod, na.rm = TRUE),
    temp_strat_mean = mean(temp_strat, na.rm = TRUE),
    aod_sd = mean(aod, na.rm = TRUE),
    temp_strat_sd = mean(temp_strat, na.rm = TRUE),
    .groups = "drop"
  )
```

Add means and center variables to prepare values to be used in the ESN:

```{r}
merra2 <-
  left_join(merra2_train_test, merra2_spatial_means_and_sds, by = c("lon", "lat")) %>%
  mutate(
    aod_stdzd = (aod - aod_mean) / aod_sd,
    temp_strat_stdzd = (temp_strat - temp_strat_mean) / temp_strat_sd
  ) %>%
  select(
    data,
    date,
    lon,
    lat,
    aod,
    aod_mean,
    aod_sd,
    aod_stdzd,
    temp_strat,
    temp_strat_mean,
    temp_strat_sd,
    temp_strat_stdzd
  )
```

Save cleaned data to the data folder:

```{r}
usethis::use_data(merra2, overwrite = TRUE)
```
